---
title: "Model for Analyzing Student Performance in Math"
output: html_document
---

<h3>Introduction: </h3>
Group Members: Rohit Kumar (rohitk2)
The data was taken from the UCI Machine Learning Repository. The link is https://archive.ics.uci.edu/ml/datasets/Student+Performance#. It provides information on 396 students from two Portugese secondary schools. The data includes 33 different variables, which include a variety of social, family and environmental factors and the final grade of that semester in math. Here is the complete variable list and descriptions copied from the documentation: 

<b> Attributes: </b>

1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira) \newline

2 sex - student's sex (binary: 'F' - female or 'M' - male) \newline

3 age - student's age (numeric: from 15 to 22) \newline

4 address - student's home address type (binary: 'U' - urban or 'R' - rural) \newline

5 famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3) \newline

6 Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart) \newline

7 Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 â???" 5th to 9th grade, 3 â???" secondary education or 4 â???" higher education) 

8 Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 â???" 5th to 9th grade, 3 â???" secondary education or 4 â???" higher education) 

9 Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') \newline

10 Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other') \newline

11 reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other') \newline

12 guardian - student's guardian (nominal: 'mother', 'father' or 'other') \newline

13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour) \newline

14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) \newline

15 failures - number of past class failures (numeric: n if 1<=n<3, else 4) \newline

16 schoolsup - extra educational support (binary: yes or no) \newline

17 famsup - family educational support (binary: yes or no) \newline

18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no) \newline

19 activities - extra-curricular activities (binary: yes or no) \newline

20 nursery - attended nursery school (binary: yes or no) \newline

21 higher - wants to take higher education (binary: yes or no) \newline

22 internet - Internet access at home (binary: yes or no) \newline

23 romantic - with a romantic relationship (binary: yes or no) \newline

24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent) \newline

25 freetime - free time after school (numeric: from 1 - very low to 5 - very high) \newline

26 goout - going out with friends (numeric: from 1 - very low to 5 - very high) \newline

27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high) \newline

28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high) \newline

29 health - current health status (numeric: from 1 - very bad to 5 - very good) \newline

30 absences - number of school absences (numeric: from 0 to 93) \newline

<b> Math Grades: </b>

31 G1 - first period grade (numeric: from 0 to 20) 

32 G2 - second period grade (numeric: from 0 to 20) 

33 G3 - final grade (numeric: from 0 to 20, output target)

Out of the 33 variables, I did not consider G1, failure, and G2 as predictors, because it would be predicting final results with prior results. I also was not interested in absences, guardian, reason for attending school, age and school. In this project I had 2 goals. 

As a student, I found this dataset pretty interesting. I am interested in how various factors affect student performance including family background, parent income and how students spend their time outside of school (including partying, drinking, dating, etc.) effect student performance in mathematics. Because of my limited ability to interact with too many people outside my demographic, it is tough to really analyze how these factors affect student performance. However, given an actual dataset, I can analyze the trends of these variables. The main goal of my project will be to find the best model to explain NOT PREDICT student performance in math, given all these factors. 

While AIC is a good tool in eliminating short, inaccurate models, I wanted a model as short as possible, even if that meant a lot more residual. In the method section, I used AIC as a tool to generate smaller but relatively accurate models. Because 31 variables is too large of a model to begin with, I binned the variables into categories, in which I analyzed the correlation these variables had with eachother, to account for collinearity and correlation with grades. Using this tool I created a subset of variables, in which I used to create more models. 

<h3> Methods: </h3>


```{r}
library(faraway)
math = read.csv("refined_math.csv")
all_variables = lm(G3 ~ . - G1 - G2 - absences - failures - guardian - reason - school - age, data = math)
summary(all_variables)
```

```{r}
#Bin 1: Leisure
cor(cbind(walc = math$Walc, dalc = math$Dalc, goout = math$goout, health = math$health, freetime = math$freetime, romantic = math$romantic, grade = math$G3))
```
Both going out (goout) and being in a romantic relationship (romantic) tend to have decently strong enough correlation with grade, while other variables do not have a strong enough correlation with grades, so I decided to eliminate other variables. Furthermore, going out and being in a romantic relationship have a correlation of only about 0.01607, which means they do not have a high collinearity.
```{r}
#Bin 2: Family Structure Category
cor(cbind(fjob = math$Fjob, mjob = math$Mjob, pstatus = math$Pstatus, famrel = math$famrel, famsup = math$famsup, fsize = math$famsize, higher = math$higher, nursery = math$nursery, fedu = math$Fedu, medu = math$Medu, grade = math$G3))
```
The variables with strong enough correlation to grades are the mother's job (mjob), the desire to pursue higher education (higher), the mother's education (medu) and father's education (fedu). However, the mother's job and mother's education have high correlation, implying collinearity. As a result, since the mother's education is more correlated with grades, I eliminated the mother's job.

```{r}
#Bin 3: Activity Outside School
cor(cbind(schoolsup = math$schoolsup, famsup = math$famsup, studytime = math$studytime, higher = math$higher, paid = math$paid, activities = math$activities, nursery = math$nursery, internet = math$internet, freetime = math$freetime, goout = math$goout, grade = math$G3))
```
The variables with strong enough correlation to grade are time spent studying (studytime), desire to pursue higher education (higher), extra paid math courses (paid), access to internet (internet) and going out (goout). I eliminated the other variables.

```{r}
#Bin 3: Location
cor(cbind(traveltime = math$traveltime, rural_or_urban = math$address, grade = math$G3))
```
While traveltime and whether a student lives in an urban or rural area have a moderate correlation implying some collinearity, both seem to have about the same amount of correlation with grades. Since I cannot conclude whether having traveltime or being in a rural area have stronger a strong correlation with grades and traveltime has only a small amount of higher correlation with grade. To be safe I will include both in my list of variables.   


After analyzing these bins, here is my list of variables: goout, romantic, higher, medu, fedu, studytime, paid, internet, address, and traveltime.

```{r}
#CHECKING VARIATIONS OF MODEL SELECTION FOR THE LINEAR MODEL BEGINNING WITH COMPLETE LIST OF VARIABLES
#---------------------------------------------------------------
#Backwards AIC Model
linear_model_template = all_variables
all_variables_back_aic = step(linear_model_template, direction = "backward", trace = 0)
```
Note: When evaluating the forward and bidirectional model, they came out to be the null model, so I did not include them.
```{r}
#CHECKING VARIATIONS OF MODEL SELECTION FOR THE LINEAR MODEL BEGINNING WITH MY OWN LIST OF VARIABLES
#---------------------------------------------------------------
#Model Containing All My Variables
my_variables = lm(G3 ~ goout + romantic + higher + Medu + Fedu + studytime + paid + internet + traveltime + address, data = math)
#Backwards AIC Model
linear_model_template = my_variables
my_variables_back_aic = step(linear_model_template, direction = "backward", trace = 0)
```
Note: When evaluating the forward and bidirectional model, they came out to be the exact same as the backwards model, so I did not include them.
```{r}
#CHECKING VARIATIONS OF MODEL SELECTION FOR THE LINEAR MODEL BEGINNING WITH MY OWN LIST OF VARIABLES WITH INTERACTION TERMS
#---------------------------------------------------------------
#Backwards AIC Model
linear_model_template = lm(G3 ~ (goout + romantic + higher + Medu + Fedu + studytime + paid + internet + traveltime + address)^2, data = math)
my_variables_interaction_back_aic = step(linear_model_template, direction = "backward", trace = 0)
#Forward AIC Model
my_variables_interaction_forward_aic = step(
  lm(G3 ~ 1, data = math), 
  scope = G3 ~ (goout + romantic + higher + Medu + Fedu + studytime + paid + internet + traveltime + address)^2, 
  direction = "forward", trace = 0)
```
Note: When evaluating the bidirectional model, it came out to be the same as the forward model, so I did not include it.

```{r}
#AIC,RSS,Parameter Table for all the 6 Models
#---------------------------------------------------------------
Number_of_parameters = c(extractAIC(all_variables)[1], extractAIC(all_variables_back_aic)[1], extractAIC(my_variables)[1], extractAIC(my_variables_back_aic)[1], extractAIC(my_variables_interaction_back_aic)[1], extractAIC(my_variables_interaction_forward_aic)[1])
AIC = c(extractAIC(all_variables)[2], extractAIC(all_variables_back_aic)[2], extractAIC(my_variables)[2], extractAIC(my_variables_back_aic)[2], extractAIC(my_variables_interaction_back_aic)[2], extractAIC(my_variables_interaction_forward_aic)[2])
RSS = c(sum(resid(all_variables)^2), sum(resid(all_variables_back_aic)^2), sum(resid(my_variables)^2), sum(resid(my_variables_back_aic)^2), sum(resid(my_variables_interaction_back_aic)^2), sum(resid(my_variables_interaction_forward_aic)^2))
models = c("ALL-1", "ALL Back AIC-2", "Linear My Var-3", "Back AIC Linear My Var-4", "Back AIC (2-way Interaction)-5", "Forward AIC (2-way Interaction)-6")
result = data.frame(models, AIC, RSS, Number_of_parameters)
colnames(result) = c("Model", "AIC", "RSS", "Coefficient #")
knitr::kable(result)
```

Out of the 6 models, I did not adhere strictly to AIC. While I was interested in RSS, I do not need a model that has the most accuracy. I wanted to provide a simple, concise explanation, that people would remember, eliminating too many interaction terms and variables. As a result I chose Model 4 despite it having a high AIC and RSS.

<h3> Results </h3>
The model I chose was:
```{r}
summary(my_variables_back_aic)
```

<h3> Discussion </h3>
While this is by no means a comprehensive guide model, it provides a small simplified set of important factors that can explain higher grades the best. My model, giving all the variables a linear relationship is easy to explain. According to my model, the key factors that explain higher grades in math are (a) going out (negavitve correlations), (b) being in a romantic relationship (negative correlation), (c) the desire to pursue a higher education (positive correlation), the mother having a higher level of education (positive correlation) and access to internet (positive correlation).

Out of curiosity my side goal was to see how all of the different variables in the data actually correlate with final grade. As a result, I analyzed the summary of the linear model consisting of all variables. Here is what I found.

<h5>Factors Correlating with Better Final Grade in Math</h5>
<ul>
  <li>male</li>
  <li>Urban</li>
  <li>family Size $\geq$ 3 </li>
  <li>mother and father having better education </li>
  <li>mother and father having a job in healthcare </li>
  <li>Parents being divorced </li>
  <li>father being teacher, but not Mother </li>
  <li>mother and father working in healthcare </li>
  <li>mother working in service but father not </li>
  <li>having less travel time to school </li>
  <li>studying longer hours (obvious) </li>
  <li>not having extra family or school support </li>
  <li>having extra paid classes </li>
  <li>not doing extra-curricular activities </li>
  <li>not attending nursery </li>
  <li>desire to pursue higher education </li>
  <li>access to internet</li>
  <li>not being in a romantic relationship </li>
  <li>avoiding daily alchol, going out, or bad health </li>
  <li>consuming weekend alchohol (shocker this had a positive correlation with better math grades)</li>
  <li>more free time </li>
  <li>better family relationship </li>
</ul> 

<h3> Appendix </h3>


The R code I used to evaluate every variable.

```{r}
summary(all_variables)
```